
---
title: "PADP8120 Final Fall 2015"
author: "Prisca Lee"
output:
  html_document:
    highlight: pygments
    theme: cerulean
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
---


# Final Exam

**“I affirm that I have received no support in taking this exam from any individual other than the Professor for this course”.PL**

**due by 5pm on Thursday, December 10th, 2015**

## Instructions

A. Fork the [Final repository](https://github.com/PADP8120-Fall2015/Final) from the course github page to your github account (i.e., just as you would with a homework assignment)

B. In RStudio, create a new project and link it to your newly forked github repo. 

C. Resave the `Final_Fall2015.Rmd` file as `Final_Fall2015_FirstInitialLastName.Rmd` (e.g., Final_Fall2015_TScott.Rmd`)

D. Complete the final within your `Final_Fall2015_FirstInitialLastName.Rmd` file. 

E. Make sure your final document renders as an `.html` file. 

F. Please **email** to me all of the materials necessary for another person to run your R Markdown file, including:
  - The R project (`.Rproj`) file
	- The R Markdown document (`.Rmd`) of your analyses
	- An HTML document (`.html`) compiled from your R Markdown document.
	- Any data or other files neede to run the analyses in your R Markdown document.

## Guidelines 

i. You may use any texts, help files, internet resources, web tutorials, etc. that you wish, but you may only discuss this exam (or receive any kind of personal assistance) from the Professor for this course. This includes a prohibition on receiving R help from any individual aside from the Professor in relation to this exam. To be clear, discussing this exam in any way with anyone besides the Professor for this course constitutes academic dishonesty. Your exam must have the following statement appended to the top (you may cut and paste it) and initialed. Exams without this statement will not be graded.



ii. Your exam write-up should be clear and legible, with answers clearly indicated and work shown. Your exam must be produced in html (.html) document produced using R Markdown (i.e., an .Rmd file). Submit both your .html file and the .Rmd file used to generate it via github to the course midterm repo. If you are having trouble accomplishing this, please refer to the [guide](http://spia.uga.edu/faculty_pages/tyler.scott/teaching/PADP8120_Fall2015/Homeworks/submitting_homework.shtml). 

iii. The exam will be graded out of 100 points. Each problem is worth a designated number of points (shown below). Partial credit *may* by given for incorrect answers if I am able to see the process by which you went wrong (and thus see what you also did correctly), so it is to your advantage to show your work. Be sure to only include relevant information, as incorrect statements appended to correct responses will result in a penalty.

iv. Your exam is **due by 5pm on Thursday, December 10th, 2015**; late exams will not be accepted. 

v. I will address clarification questions until 8pm on Wednesday, December 9th. After that time, you are expected to complete the exam without any assistance.  

vi. Please contact me if you have any questions or concerns.


# Problems 

## Set 1
The first set of questions pertain to Olympic gold medal-winning 100 meter sprint times for men and women.
```{r eval=TRUE}
sprinters <- read.csv("input/sprinters.csv")
```

| Variable | Description |
|:--|:--|
| `year` |    Year of race |
| `finish` |     completion time in seconds |
| `women` |    indicator for female or male time |

(@) (4 points) Estimate the model
    $$
    \begin{aligned}[t]
    \mathtt{finish}_i &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{women}_i \times \mathtt{year}_i  + \epsilon_i
    \end{aligned}
    $$
    
```{r}
mod1 <- lm(finish~year + women +women*year, data = sprinters)
summary(mod1)
```

(@) (4 points) Given your results for the model above, what is the marginal effect of `year` for women? For men? Interpret each in a sentence. 

The marginal effect of year for women is -0.006, meaning there is a negative relationship between the finish time and year for women. The marginal effect of year for men is -0.011, meaning there is a also a negative relationship between the finish time and year for men. 


(@) (4 points) Create a plot of the fitted values (with confidence intervals) with respect to `year`, grouped by `women`. Do not use `geom_smooth()`.
```{r}
library(ggplot2)
```

```{r}
pred2_df <- expand.grid(year = seq(1900, 2600, by = 100), women = c(0,1))


ggplot(bind_cols(pred2_df, as.data.frame(pred2)), aes(x = year, y = fit, ymin = lwr, ymax = upr, colour = factor(women), fill = factor(women))) + geom_ribbon(alpha = 0.2, colour = NA) + geom_line()pred2.df <- expand.grid(year = seq(1900, 2600, by = 100), women = c(0,1))
pred2 <- predict(mod1, newdata = pred2_df, interval = "confidence")
ggplot(bind_cols(pred2_df, as.data.frame(pred2)), aes(x = year, y = fit, ymin = lwr, ymax = upr, colour = factor(women), fill = factor(women))) + geom_ribbon(alpha = 0.2, colour = NA) + geom_line()
```


(@) (2 points) Create the same plot as in the previous problem for the model
    $$
    \begin{aligned}[t]
    \log (\mathtt{finish}_i) &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{year}_i \times \mathtt{women}_i + \epsilon_i
    \end{aligned}
    $$
    
```{r}
mod2 <- lm(log(finish)~year + women + year*women, data = sprinters)

pred3 <- predict(mod2, newdata = pred2_df, interval = "confidence")
ggplot(bind_cols(pred2_df, as.data.frame(pred3)), aes(x = year, y = fit, ymin = lwr, ymax = upr, colour = factor(women), fill = factor(women))) + geom_ribbon(alpha = 0.2, colour = NA) + geom_line()
```

(@) (4 points) Explain in no more than 2 sentences how this specification differs from the prior specification. This reduces the heteroskedasticity, and resultantly, the confidence intervals change.  

This model specification differs from the previous one in that the dependent variable of finish- completion time in seconds- has been transformed by log. 

(@) (6 points) Rerun the analysis and recreate the plot for the model:
    $$
    \begin{aligned}[t]
    \mathtt{Finish}_i &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{year}_i × \mathtt{women}_i \\
	& + \beta_4 \mathtt{year}_i^2 + \beta_5 \mathtt{year}_i^2 × \mathtt{women}_i + \epsilon_i
    \end{aligned}
    $$

```{r}
mod3 <- lm(finish~year + women + year*women + I(year^2) + I(year^2)*women, data = sprinters)
summary(mod3)

pred4 <- predict(mod3, newdata = pred2_df, interval = "confidence")
ggplot(bind_cols(pred2_df, as.data.frame(pred4)), aes(x = year, y = fit, ymin = lwr, ymax = upr, colour = factor(women), fill = factor(women))) + geom_ribbon(alpha = 0.2, colour = NA) + geom_line()
```

(@) (4 points) Explain in 2-3 sentences how this specification differs from the previous models. 

This model has the additional interaction terms of year^2 and year^2xwomen. The interaction term present in all three models of yearxwomen takes into account the potential relationship between the year of the Olympics and sex. The additional year^2 and year^2xwomen corrects for some of the issues with residuals in year and then creates a new interaction term for year^2 and its relationship to sex.

(@) (4 points) Compare the visual fit of these models to the data within the observed period. Which (if any) do you find to be plausible fits?

```{r}
plot(sprinters$finish~mod1$fitted.values)
abline(0,1)
plot(sprinters$finish~mod2$fitted.values)
abline(0,1)
plot(sprinters$finish~mod3$fitted.values)
abline(0,1)
```

Visually, it appears that all three do a similar job of fitting the observed data. However, it appears that model 3 may be slightly better at fitting predicted values with observations.

(@) (6 points) Do these models have different predictions for the Olympics of 2156? Why or why not?

These models have different predictions for the Olympics of 2156 because nonlinear transformations have been made to the second and third models, thus changing the inear relationship and correlation between variables. This ultimately leads to different predicitions for a given year.

(@) (6 points) Now create a new variable, the ratio of men’s time to women’s time in each year. Log-transform this variable and regress it on year. Plot the results, with confidence intervals, on the scale of the ratio men’s time to women’s time (i.e., transform it back from the natural logarithm).

##Set 2

For this problem, we will use a cleaned-up version of the dataset of Michael Ross, "Does Oil Hinder Democracy?" *World Politics*, 2001.That paper estimated a time series cross-section model of Polity scores regressed on oil exports and a battery of controls. In this problem, we will focus on a single cross-section, and instead focus on model fitting.

Load and pre-process this data using the following code:
```{r message = FALSE}
library(dplyr)
rossoil <- read.csv("input/rossoildata.csv") %>%
   arrange(id1, year) %>%
   group_by(id1) %>%
   mutate(oilL5 = lag(wdr123, 5) / 100,
          metalL5 = lag(wdr313, 5) / 100,
          GDPpcL5 = lag(wdr135, 5) / 100,
          islam = islam / 100) %>%
   filter(year == 1990) %>%
   select(regime1, oilL5, metalL5, GDPpcL5, islam, oecd, cty_name, id, id1) %>%
   na.omit() %>%
   ungroup()
```

The above code:

- `lag()` calculates the lag values of `oil`, `metal`, and `GDPcap`. The second argument of `5`, means that it calculates a lag of 5.[^1]
- `group_by` ensures that the lag values are only calculated within each country.
- Keeps only observations from 1995 and a subset of variables
- Omits missing values with `na.omit()`.
- `ungroup()` ensures that the data is no longer grouped by `id`.
  If you try to use `summarize()` while the data is still grouped by `id`, you would
  not get the results that you thought you would.

The variables are: 

| Variable | Description |
|:--|:--|
| `regime1` |    1–10 scale increasing in democracy; computed from Polity components |
| `oilL5` |      Fuel exports as a proportion of GDP, lagged 5 years |
| `metalL5` |    Ore and mineral exports as a proportion of GDP, lagged 5 years |
| `GDPpcL5` |    per capita GDP in PPP dollars, lagged 5 years |
| `islam` |      Muslims as a proportion of population, 1970 data |
| `oecd` |       Dummy for rich industrialized countries |
| `cty_name` |   The name of the country observed |
| `id` |         A three character abbreviation of the country name |
| `id1` |        A numeric country code |


(@) (4 points) Estimate a linear regression of `regime1` on `oilL5`, `metalL5`, `GDPpcL5`, `islam`, and `oecd`.

```{r}
ross.mod1 <- lm(regime1~oilL5 + metalL5 + GDPpcL5 + islam + oecd, data = rossoil)
summary(ross.mod1)
```

(@) (8 points) Provide a substantive interpetation of each coefficient and note whether a given coefficient is statistically significant. 

Each additional unit of fuel exports as a proportion of GDP, lagged 5 years predicts a 2.248 decrease in units of regime. Each additional unit of ore and mineral exports as a proportion of GDP, lagged 5 years predicts a 0.072 decrease in units of  regime; this is statistically significant at the 0.1 level. Each additional unit of per capita GDP in PPP dollars, lagged 5 years predicts a 0.009 increase in units of regime. Each additional unit of Muslims as a proportion of population, using 1970 data predicts a 4.442 decrease in units of  regime; this is statistically significant at the highest level of 0. Each additional oecd unit predicts a 1.845 increase in units of regime. 

(@) (4 points) Calculate the expected change in `regime1` given a change in `oilL5` from the 50th percentile to the 95th percentile of the fully observed data, all else equal.

(@) (8 points) Using the residuals from the regression in above, create the following diagnostic plots:
    
    - Plot the standardized residuals against the fitted values
    
```{r}
ross.mod1.stdres <- rstandard(ross.mod1)

plot(ross.mod1$fitted.values, ross.mod1.stdres)
abline (0,0)
```
  
    - Plot the standardized residuals against each covariate
    
```{r}
plot(ross.mod1.stdres~rossoil$oilL5)
plot(ross.mod1.stdres~rossoil$metalL5)
plot(ross.mod1.stdres~rossoil$GDPpcL5)
plot(ross.mod1.stdres~rossoil$islam)
plot(ross.mod1.stdres~rossoil$oecd)
```
  
    - Plot $Y$ versus $\hat{Y}$
    
```{r}
plot(rossoil$regime1~ross.mod1$fitted.values)
abline(0,1)
```

(@) (6 points) What do these diagnostics tell you about the presence of heteroskedasticity, specification error, and outliers?

The plot of standardized residuals against fitted values along with the plots of standardized residuals against covariates indicates that there is heteroskedasticity. The plot of observed values of fitted values indicates that there may be some model specification issues; the points do not create a linear structure. There does not appear to be any outliers.

(@) (4 points) Rerun the regression using any 2 non-linear transformations for any of the covariates ("x variables") based upon your theoretical expectations and/or your diagnostics above). You're welcome to explore several specifications, but only present the specification that you think is best and describe why you choose to transform what you did. You WILL NOT be graded on whether the model is ultimatly a good fitting model.

```{r}
ross.mod2 <- lm(regime1~ I(oilL5^2) + metalL5 + log(GDPpcL5) + islam + oecd, data = rossoil)
summary(ross.mod2)
```

In my new specification, I choose to exponentiate oil because of its residual plot above, and theoretically I believe that the effect of oil may be nonlinear. For my second transformation, I choose to transform GDP per capita. lagged 5 years by log because the data for this variable is skewed. 

(@) (4 points) For your new model, record and compare the adjusted $R^2$, AIC score, and BIC score for your new model.

```{r}
summary(ross.mod1)
summary(ross.mod2)
AIC(ross.mod1, ross.mod2)
BIC(ross.mod1, ross.mod2)
```

The adjusted r-squared is 0.471 for the original specification and 0.552 for the new speicifcation with transformations. The AIC 341.546 for the new model and 329.856 for the second model. The BIC is 357.285 for the first model and 345.596 for the second model. All three measures indicate that the second model is better specified than the first one.

(@) (4 points) For both of your new specifications, compute the expected change in `regime1` given a change in `oilL5` from the 50th percentile to the 95th percentile of the fully observed data.

(@) (4 points) Assess how much substantive difference finding the best model makes. Be specific and concrete; i.e., show what each model does. 

Finding the best model includes having the right transformations- whether linear or nonlinear, deciding which variables to include, and not violating the key assumptions. Substantively, this makes a significant difference. In the model above, it corrects a bit for heteroskedasticity by reducing residual variances, and predicted values fit observed values better than the first model.
  
(@) (4 points) What other problems in the specification or estimation method remain unaddressed by our efforts (no more than 2-3 sentences)?

The issue of heteroskedasticity still needs to be further reduced, and the distribution of the data is uneven. 

(@) Try a new specification that alters the dependent variable in order to address the heteroskedasticity problem. Assess whether your new model successfully addresses the problem. 

```{r}
ross.mod3 <- lm(sqrt(regime1)~I(oilL5^2) + metalL5 + log(GDPpcL5) + islam + oecd, data = rossoil)
summary(ross.mod3)
ross.mod3.stdres <- rstandard(ross.mod3)
plot(ross.mod3.stdres~ross.mod3$fitted.values)
abline(0,0)
```

The new model for in which the square root was taken for the dependent vairable of regime1 does not seem to have significantly improved heteroskedasticity.

